---
title: "Getting started with dendroMixR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dendroMixR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6.5,
  fig.height = 4.5,
  dpi = 110
)
set.seed(1)
```

## The problem this package solves

Fitting a finite mixture model requires choosing the number of components
$k$. In practice $k$ is unknown, so it is common to *overfit* on purpose:
fit with more components than you believe are really there, then try to
make sense of the result. Unfortunately, overfitted mixture models come
with a real statistical cost. Parameter estimates for an overfitted mixing
measure converge much more slowly than for a correctly-specified one —
as slowly as $n^{-1/8}$ for some Gaussian mixtures — and the fit typically
contains redundant, near-duplicate components that are hard to interpret.

`dendroMixR` implements the method from Do, Do, McKinley, Terhorst, and
Nguyen, *"Dendrogram of mixing measures: Hierarchical clustering and model
selection for finite mixture models"* (Biometrika, 2025). The idea is to
take a single overfitted fit and repeatedly merge its two most similar
components (its "atoms"), producing a nested sequence of mixing measures
with $k, k-1, \dots, 1$ components — much like agglomerative hierarchical
clustering, but applied to mixture atoms rather than data points. The
paper shows that this merging procedure recovers the fast, root-$n$
convergence rate at the true number of components, even though the
starting overfitted fit converged slowly, and it does so without ever
re-fitting the model.

This vignette shows the two core functions in action:

* `dendrogram_mixing()` performs the merging and returns a hierarchical
  clustering object (`hc`), the sequence of intermediate mixing measures
  (`Gs`), and the merge history (`merge_pair`).
* `plot_dendrogram_mixing()` visualizes the result as a dendrogram, with
  one parameter dimension shown at a time.

```{r setup}
library(dendroMixR)
```

## A minimal worked example

Before working with simulated data, it helps to see `dendrogram_mixing()`
on a small, hand-built mixing measure so the mechanics are clear. Here are
six atoms in one dimension, arranged so that atoms 1-2, 3-4, and 5-6 are
each close together (as if an EM fit had produced redundant near-duplicate
components near three "true" locations at roughly $-2$, $0$, and $3$):

```{r toy-example}
ps_toy     <- c(0.05, 0.30, 0.05, 0.30, 0.15, 0.15)
thetas_toy <- c(-2.1, -1.9, 0.05, -0.05, 3.0, 3.2)

dmm_toy <- dendrogram_mixing(ps_toy, thetas_toy)
str(dmm_toy, max.level = 1)
```

`dmm_toy$hc$height` gives the dissimilarity at which each merge happened.
Because atoms 1-2, 3-4, and 5-6 are near-duplicates, we expect the first
few merges to happen at a much smaller height than the later ones that
join genuinely distinct components:

```{r toy-heights}
dmm_toy$hc$height
```

The heights jump by roughly two orders of magnitude between the third and
fourth merges — exactly the pattern we'd want, since the first three
merges clean up redundant atoms while the last two join real, separated
components.

## Example 1: strongly identifiable mixtures (known variance)

When the component variance is known and shared across components (the
*strongly identifiable* case), `dendrogram_mixing()` only needs weights
and means: `sigmas = NULL`. This case also covers the common practice of
overfitting via a hard-clustering method such as $k$-means, which
implicitly assumes equal, fixed variance across clusters.

We simulate data from a well-separated 3-component 1-D Gaussian mixture
with unit variance, then deliberately overfit with $k = 8$ clusters:

```{r strong-sim}
n   <- 2000
ps0 <- c(0.3, 0.4, 0.3)
mu0 <- c(-4, 0, 5)
k0  <- 3

z <- sample(k0, n, replace = TRUE, prob = ps0)
x  <- rnorm(n, mu0[z], sd = 1)

k_over <- 8
km <- kmeans(x, centers = k_over, nstart = 20)
ps_over    <- as.numeric(table(km$cluster)) / n
theta_over <- as.numeric(km$centers)
```

Now merge the overfitted fit down with `dendrogram_mixing()`:

```{r strong-merge}
dmm1 <- dendrogram_mixing(ps_over, theta_over)
dmm1$hc$height
```

`dmm1$Gs` holds the mixing measure at every level of the tree, from
$k = 8$ atoms down to $k = 1$. Cutting at the true $k_0 = 3$
(`Gs[[k_over - k0 + 1]]`) recovers weights and means very close to the
truth (0.3/0.4/0.3 and $-4, 0, 5$), even though no re-fitting took place:

```{r strong-merged}
merged3 <- dmm1$Gs[[k_over - k0 + 1]]
merged3$ps
merged3$thetas
```

`plot_dendrogram_mixing()` visualizes the merge sequence. Each column of
points is one level of the tree ($k$ atoms down to 1); tracing the gray
branches from left to right shows which atoms combine and at what height:

```{r strong-plot}
plot_dendrogram_mixing(dmm1, main = "Strong identifiability: 1D merge")
```

## Example 2: weakly identifiable mixtures (location-scale Gaussian)

Location-scale Gaussian mixtures — where both the mean *and* the
covariance are estimated — are *weakly identifiable* and are known to
have especially slow convergence when overfitted. Pass a `sigmas`
argument (a list of covariance matrices for multivariate data, or a
numeric vector of variances in 1-D) to use the weak-identifiability merge
rule.

We simulate 2-D data from three Gaussians with different covariance
structures and fit an overfitted $k=8$ Gaussian mixture with `mclust`:

```{r weak-sim}
library(mclust)

n <- 1500
ps0  <- c(1/3, 1/3, 1/3)
mus0 <- matrix(c( 2, 1,
                   0, 6,
                  -2, 1), nrow = 3, byrow = TRUE)
sigmas0 <- list(
  matrix(c(0.50,  0.30,  0.30, 1.00), 2, 2),
  matrix(c(0.50, -0.10, -0.10, 0.10), 2, 2),
  matrix(c(0.25,  0.30,  0.30, 2.00), 2, 2)
)
k0 <- 3

z <- sample(k0, n, replace = TRUE, prob = ps0)
x <- matrix(0, n, 2)
for (i in seq_len(n)) x[i, ] <- MASS::mvrnorm(1, mus0[z[i], ], sigmas0[[z[i]]])

k_over <- 8
fit <- Mclust(x, G = k_over, modelNames = "VVV", verbose = FALSE)

ps_over     <- fit$parameters$pro
mus_over    <- t(fit$parameters$mean)
sigmas_over <- lapply(seq_len(k_over), function(j) fit$parameters$variance$sigma[, , j])
```

```{r weak-merge}
dmm2 <- dendrogram_mixing(ps_over, mus_over, sigmas_over)
dmm2$hc$height
```

Cutting at $k_0 = 3$ again recovers parameters close to the truth:

```{r weak-merged}
merged3_weak <- dmm2$Gs[[k_over - k0 + 1]]
merged3_weak$ps
matrix(merged3_weak$thetas, ncol = 2)
```

`dendrogram_mixing()` returns a standard `hclust` object in `$hc`, so
familiar tools like `cutree()` work directly on it — here recovering
which of the 8 overfitted components belong to which of the 3 underlying
clusters:

```{r weak-cutree}
cutree(dmm2$hc, k = 3)
```

Because parameters are multivariate here, `plot_dendrogram_mixing()`
takes a `dim` argument to choose which coordinate to display:

```{r weak-plot}
plot_dendrogram_mixing(dmm2, dim = 1, main = "Weak identifiability: dim 1")
```

```{r weak-plot-2}
plot_dendrogram_mixing(dmm2, dim = 2, main = "Weak identifiability: dim 2")
```

## Choosing where to cut the dendrogram

Both examples above cut the tree at the known true $k_0$ for illustration.
In practice $k_0$ is unknown, and the heights in `dmm$hc$height` are
themselves informative: the paper shows that merges among redundant,
overfitted atoms happen at heights shrinking to 0 as the sample size
grows, while merges between genuinely distinct components stay bounded
away from 0 — visible above as the sharp jump in height between the
early and late merges in both examples. A more formal, likelihood-aware
selection rule (the paper's Dendrogram Selection Criterion, compared
there against AIC/BIC/ICL) will be covered in a follow-up vignette on
model selection.

## Session info

```{r session-info}
sessionInfo()
```
